---
title: "Lab 3 - Colleges"
author: "Dr Bob"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(knitr)
load("colleges2.rda")
data(colleges2)
```


## About the data

The College Scorecard (<https://collegescorecard.ed.gov/>) is a tool from the U.S. Department of Education for exploring information about colleges and universities. The complete data set can be downloaded and has 6,806 rows and 1,986 columns. The derived dataframe `colleges2` has data on 73 variables for 1732 institutions that meet these criteria:

* Is a Title IV school (processes federal finanical aid)
* The "Predominant undergraduate degree awarded" is at least a bachelor's degree

In this lab you'll perform descriptive analysis to explore these data. 

## Instructions. 

Use the corresponding R script "Lab 3 Colleges.R" to learn how to use R to perform some descriptive statistics. When you arrive at the report-ready output, paste that code into the appropriate code chunks below. 


## 1. Summary Statistics for Some Key Variables

First we'll examine summary statistics for a small number of quantitative variables.

```{r summstats}
library(skimr, include=FALSE)
skim(colleges2, UGDS, netprice, SAT_AVG, GRAD_DEBT_MDN, FAMINC, MD_EARN_WNE_P10
     , ADM_RATE, RET_FT4, PCTPELL, PCTFLOAN, FIRST_GEN, GT_28K_P10)
```
  
  
# Explore Colleges & Universities Throughout the U.S.


### 2. Histogram of Average Net Price. 

Here you'll start with base code for making a histogram, which is provided in the R Script. Then I'll suggest some improvements. As you make each change and see the result, in the sentences below, replace the bolded words in the sentences below with an answer about what changed.

```{r netprice-hist}

```

1. After adding `bindwidth=2500` to the geom_histogram() function, the bins got **ANSWER**.
2. After adding `color="black"`, the **ANSWERs** are now black.
3. After adding `fill="lightblue`, the **ANSWERs** are now light blue.


### 3. Net Price vs Type of Institution

The variable *control* indicates whether a school is public or private. Produce comparative histograms for the two levels of that variable. We'll see two ways to do this. 

#### 3a. Version 1: use colors to show the two histograms on the same plot space.

```{r price-control-color}

```

#### 3b. Version 2: use panels

Another way to show two histograms is through the use of panels, which is referred to as a *facet* in the ggplot2 *grammar of graphics*. 

```{r pricepanels}

```


## 4. Boxplot of number of degree-seeking undergraduate students by type of institution

```{r ugds-box}

```

If you're like me, you're probably wondering what the deal is with that big-time outlier for private schools. We'll learn a little about isolating data the tidyverse way and then find out what's the deal. 


## 5. List the 10 Largest Private Institutions by number of undergraduates

### Data Manipulation the Tidyverse way

We often must extract data from a table before we can do what we want. The package DPLYR is one of the packages that gets loaded when we run `library(tidyverse)`. DPLYR is used for data manipulation. Go through the series of commands in the R script one at a time, reading the provided comments, to get introduced to the fundamentals of this important package. 

```{r ten-largest}

```

What is the largest institution? This will probably surprise you (it did me). Do a Google search (sorry Bing) for that university and scroll down to a 2019 article from insidehighered.com. Why is it so big? (Replace ANSWER below with one complete sentence.)

Which school and why: **ANSWER.**



# Explore Colleges & Universities in Michigan

## 6. Create a data frame of MI schools

Now apply some of the data manipulation techniques you learned above to create a reduced table named *MIcolleges*. Follow through the provided comments to arrive at command that results in a table with 43 observations and 7 columns. Paste it here above the provided  kable (Knit Table) command, which will display the data frame in a nicer looking table. Uncomment that kable command.

```{r mi-colleges}

#kable(MIcolleges)
```


## 7. Histogram of Average Net Price with a reference line for GVSU 

Follow the guidance provided in the R script. Then paste the code below. 

```{r netprice-mich-hist}

```


## 8. Explore Net Price by Control - boxplots and beyond

Create a "violin plot" of net price

```{r netprice-violin}

```

## 9. ACT Combined Midpoint

Create a histogram and a violin plot showing quartiles of ACTCMMID. Both should have a reference line for GVSU. 

```{r ACTCMMID}

```

## 10. Plot three more variables

Create violin plots for the following variables

|Variable|Description|
|:-|:-----|
|RET_FT4 | First-time, full-time student retention rate at four-year institutions|
|FIRST_GEN | Percentage first-generation students|
|GT_28K_P10 | Share of students earning over $28,000/year (threshold earnings) 10 years after entry|

```{r three-violins}

```

